---
title: "STAT4610_HW3"
author: "John Dennis"
format: pdf
---

```{r,message=FALSE,echo=FALSE}
library(ISLR2)
library(tidyverse)
library(ggplot2)
```

### Problem 1

#### Part A

To find the prob that $x^{\ast 1}$ is *not* $x_1$ is simply the complement that it is. Because there are n samples, that means that there's a $\frac{1}{n}$ chance that it's the same. That means the complement is just that probability subtracted from one or:

$$
1-\frac{1}{n}=\boxed{\frac{n-1}{n}}
$$

#### Part B

To find the prob that neither $x^{\ast 1}$ nor $x^{\ast 2}$ include $x_1$ we must take the previous problem and just add on. The first prob is $\frac{n-1}{n}$. And since each draw is independent, the probability that $x^{\ast 2}$ is not $x_1$ is also $\frac{n-1}{n}$. Multiplying the two and we get:

$$
\bigg(1-\frac{1}{n}\bigg) \times \bigg(1-\frac{1}{n}\bigg) = \boxed{\bigg(1-\frac{1}{n}\bigg)^2}
$$

#### Part C

Again, just another extension. For a sample size of $n$ we simply multiply the probability we already found in A by itself $n$ times:

$$
\boxed{\bigg(1-\frac{1}{n}\bigg)^n}
$$

#### Part D

If we evaluate $\lim_{n \rightarrow \infty}$ of the probability we're building on, we'll find the prob that as $n \rightarrow \infty$ that the bootstrapped data set won't have $x_1$. So we simply take the complement of what we find:

$$
1-\lim_{n \rightarrow \infty}\bigg(1-\frac{1}{n}\bigg)^n = 1 - e^{-1} \approx 0.6321206
$$

#### Part E

```{r}
n=100
store <- rep(NA, 200)
for (i in 1:200) {
store[i] <- sum(sample(1:n, replace=TRUE) == 7) > 0
}
mean(store)
```

In this problem `mean(store)` is counting the number of times the sample contains 7 at least once out of the total sample or the probability that that 7 shows up in a sample. It is what I expected because it's essentially n=200 of our previous probability so it should start approaching $1-e^{-1}$.

#### Part F

```{r,echo=FALSE}
n_values <- 1:1000
sim_results <- numeric(1000)

set.seed(307)
for (n in n_values) {
  # Perform exactly one simulation per value of n
  sim_results[n] <- any(sample(1:n, size = n, replace = TRUE) == 1)
}

plot_data <- data.frame(
  n = n_values,
  In_Sample = as.numeric(sim_results)
)

ggplot(plot_data, aes(x = n, y = In_Sample)) +
  # Single simulation points at 0 or 1
  geom_point(alpha = 0.2, color = "red", size = 0.5) +
  # The theoretical limit you calculated in Part D
  geom_hline(yintercept = 1 - exp(-1), color = "blue", size = 1) +
  annotate("text", x = 750, y = 0.68, label = "Limit: 0.632", color = "blue") +
  labs(
    title = "Bootstrap Inclusion: Simulation vs. Theoretical Limit",
    subtitle = "Red dots = Single simulation (1 = x1 present, 0 = x1 absent)",
    x = "Sample Size (n)",
    y = "Inclusion Probability"
  ) +
  theme_bw()
```

#### Part G

$$
\begin{gathered}
Var(\bar{x}) = Var\left(\frac{1}{n}\sum_{i=1}^{n}x_i\right) \\
= \frac{1}{n^2} Var\left(\sum_{i=1}^{n}x_i\right) \\
= \frac{1}{n^2} \sum_{i=1}^{n} Var(x_i) \\
= \frac{1}{n^2} (n\sigma^2) \\
\boxed{= \frac{\sigma^2}{n}}
\end{gathered}
$$

#### Part H

$$
\begin{gathered}
Var(\bar{x}) = Var\left(\frac{1}{n}\sum_{i=1}^{n}x_i\right) \\
= \frac{1}{n^2} \left[ \sum_{i=1}^{n}Var(x_i) + \sum_{i \neq j}Cov(x_i, x_j) \right] \\
= \frac{1}{n^2} \left[ n\sigma^2 + n(n-1)\rho\sigma^2 \right] \\
= \frac{\sigma^2}{n} + \frac{n(n-1)\rho\sigma^2}{n^2} \\
\boxed{= \rho\sigma^2 + \frac{1-\rho}{n}\sigma^2}
\end{gathered}
$$

### Problem 3

```{r}
data(College)
```

This data set contains information on a large number of US College from the 1995 issue of the US News and World Report. In our case $Y$ will be `Grad.Rate`. We'll be attempting to predict using factors such as if the university is Private or not, Number of new students enrolled, Pct. of new students from the top 10% of H.S. Class, and so on. More specifically $\mathbf{X}$ will be `Private`, `Enroll`, `Top10perc`, `Top25perc`, `F.Undergrad`, `P.Undergrad`, `Room.Board`, `Books`, `Personal`, `PhD`, `Terminal`, `S.F.Ratio`, `perc.alumni`, and `Expend` (p=14).
