---
title: "STAT4610_HW3"
author: "John Dennis"
format: pdf
---

```{r,message=FALSE,echo=FALSE}
library(ISLR2)
library(ggplot2)
library(gbm)
library(BART)
library(tree)
library(dplyr)
library(randomForest)
```

### Problem 1

#### Part A

To find the prob that $x^{\ast 1}$ is *not* $x_1$ is simply the complement that it is. Because there are n samples, that means that there's a $\frac{1}{n}$ chance that it's the same. That means the complement is just that probability subtracted from one or:

$$
1-\frac{1}{n}=\boxed{\frac{n-1}{n}}
$$

#### Part B

To find the prob that neither $x^{\ast 1}$ nor $x^{\ast 2}$ include $x_1$ we must take the previous problem and just add on. The first prob is $\frac{n-1}{n}$. And since each draw is independent, the probability that $x^{\ast 2}$ is not $x_1$ is also $\frac{n-1}{n}$. Multiplying the two and we get:

$$
\bigg(1-\frac{1}{n}\bigg) \times \bigg(1-\frac{1}{n}\bigg) = \boxed{\bigg(1-\frac{1}{n}\bigg)^2}
$$

#### Part C

Again, just another extension. For a sample size of $n$ we simply multiply the probability we already found in A by itself $n$ times:

$$
\boxed{\bigg(1-\frac{1}{n}\bigg)^n}
$$

#### Part D

If we evaluate $\lim_{n \rightarrow \infty}$ of the probability we're building on, we'll find the prob that as $n \rightarrow \infty$ that the bootstrapped data set won't have $x_1$. So we simply take the complement of what we find:

$$
1-\lim_{n \rightarrow \infty}\bigg(1-\frac{1}{n}\bigg)^n = 1 - e^{-1} \approx 0.6321206
$$

#### Part E

```{r}
n=100
store <- rep(NA, 200)
for (i in 1:200) {
store[i] <- sum(sample(1:n, replace=TRUE) == 7) > 0
}
mean(store)
```

In this problem `mean(store)` is counting the number of times the sample contains 7 at least once out of the total sample or the probability that that 7 shows up in a sample. It is what I expected because it's essentially n=200 of our previous probability so it should start approaching $1-e^{-1}$.

#### Part F

```{r,echo=FALSE}
n_values <- 1:1000
sim_results <- numeric(1000)

set.seed(307)
for (n in n_values) {

  sim_results[n] <- any(sample(1:n, size = n, replace = TRUE) == 1)
}

plot_data <- data.frame(
  n = n_values,
  In_Sample = as.numeric(sim_results),
  theoretical_p = 1 - (1 - 1/n_values)^n_values
)

ggplot(plot_data, aes(x = n, y = In_Sample)) +
  
  geom_point(alpha = 0.2, color = "skyblue", size = 0.5) +

  geom_line(aes(y = theoretical_p), color = "magenta", size = 1) +
  geom_hline(yintercept = 1 - exp(-1), linetype = "dashed") +
  labs(
    title = "Probability that x1 is in the Bootstrap Sample",
    subtitle = "Simulation vs. Theoretical Curve",
    x = "Sample Size (n)",
    y = "Probability"
  ) +
  theme_bw()
```

#### Part G

$$
\begin{gathered}
Var(\bar{x}) = Var\left(\frac{1}{n}\sum_{i=1}^{n}x_i\right) \\
= \frac{1}{n^2} Var\left(\sum_{i=1}^{n}x_i\right) \\
= \frac{1}{n^2} \sum_{i=1}^{n} Var(x_i) \\
= \frac{1}{n^2} (n\sigma^2) \\
\boxed{= \frac{\sigma^2}{n}}
\end{gathered}
$$

#### Part H

$$
\begin{gathered}
Var(\bar{x}) = Var\left(\frac{1}{n}\sum_{i=1}^{n}x_i\right) \\
= \frac{1}{n^2} \left[ \sum_{i=1}^{n}Var(x_i) + \sum_{i \neq j}Cov(x_i, x_j) \right] \\
= \frac{1}{n^2} \left[ n\sigma^2 + n(n-1)\rho\sigma^2 \right] \\
= \frac{\sigma^2}{n} + \frac{n(n-1)\rho\sigma^2}{n^2} \\
\boxed{= \rho\sigma^2 + \frac{1-\rho}{n}\sigma^2}
\end{gathered}
$$

### Problem 3

```{r}
data(College)

college_clean <- College %>%
  select(Grad.Rate, Private, Enroll, Top10perc, Top25perc, 
         F.Undergrad, P.Undergrad, Room.Board, Books, 
         Personal, PhD, Terminal, S.F.Ratio, perc.alumni, Expend)
```

#### Part 1

This data set contains information on a large number of US College from the 1995 issue of the US News and World Report. In our case $Y$ will be `Grad.Rate`. We'll be attempting to predict using factors such as if the university is Private or not, Number of new students enrolled, Pct. of new students from the top 10% of H.S. Class, and so on. More specifically $\mathbf{X}$ will be `Private`, `Enroll`, `Top10perc`, `Top25perc`, `F.Undergrad`, `P.Undergrad`, `Room.Board`, `Books`, `Personal`, `PhD`, `Terminal`, `S.F.Ratio`, `perc.alumni`, and `Expend` (p=14).

#### Parts 2,3,4: Q1

The research question that we aim to answer or gain insight on is that given a certain school's demographics and institutional characteristics, can we predict the graduation rate? Not only that, but what are the most important factors to contributing to that rate. We'll be using tree based methods and Cross Validation to tune parameters to refine a model to predict the grad rate from $\mathbf{X}$

The results provide valuable information for both the institutions and prospective students. From the University's side, identifying factors that have great influence to their graduation rate will allow them to prioritize those aspects. By potentially increasing funding or resource allocation to target those factors, they may also be able to increase enrollment rates or general interest in their school.

The reason tree-based methods seem appropriate is that there likely exist non-linear relationships between the predictors and the response. Rather than attempting to use linear/multiple linear regression to model these relationships, trees will take an entirely different approach bypassing the non-linear constraint. Additionally, Ensemble Methods such as bagging and random forests will provide more robustness than a simple regression tree.

#### Parts 5,6: Q2

##### Data split

```{r,echo=FALSE}
set.seed(303)
rows <- sample(1:nrow(college_clean), size = floor(0.20*nrow(college_clean)), replace = FALSE)

validation_set <- college_clean[rows,]
CV_set <- college_clean[-rows,]
```

##### Regression Tree

I fit a regression tree using `tree` and then use `cv.tree` for the CV. I then plot the results in terms of the MSE not deviance as is the usual output for `cv.tree`.

```{r,echo=FALSE}
#Creating the tree model and seeing information about it
regTree <- tree(Grad.Rate~.,data=CV_set)
#summary(regTree)

#Performing CV on it and finding num of Nodes with lowest dev
cv_regTree <- cv.tree(regTree)
best_regTree_size <- cv_regTree$size[which.min(cv_regTree$dev)]
best_regTree_MSE <- min(cv_regTree$dev)/nrow(college_clean)

#Plotting to see how Tree size affects MSE
plot(cv_regTree$size, cv_regTree$dev/nrow(college_clean), type = "b", 
     xlab = "Tree Size (Number of Terminal Nodes)", 
     ylab = "CV MSE")

cat(sprintf("After doing Cross Validation to determine the best size of tree \n
I've found that %d is the best size with an MSE of %.3f (RMSE = %.3f)", best_regTree_size, best_regTree_MSE,sqrt(best_regTree_MSE)))
```

The plot shows that, initially, we can expect a large drop off in the MSE as the number of terminal nodes increase. As the number of terminal nodes increase, the MSE doesn't necessarily improve and this can be attributed to the fact that the model may be trying to "memorize" noise that doesn't improve it.

##### Bagging

```{r,echo=FALSE}
#The randomForest function automatically runs OOB and we set mtry to 14 because there's p predictors
bag <- randomForest(Grad.Rate~.,data=CV_set,mtry=dim(CV_set)[2]-1,importance=TRUE,ntree=500)
#We can grab the best mse using the following code
bag_oob_MSE <- bag$mse[length(bag$mse)]
#plotting for further commentary
plot(bag)
```

##### Random Forests

```{r,echo=FALSE}
nodesize.values <- seq(from = 1, to = 21, by = 5)
mtry.values <- 1:13 #13 predictors excluding the last because of the previous bagging.

oob.MSE <- matrix(NA, nrow = length(nodesize.values), ncol = length(mtry.values)) # Matrix to store OOB errors

for (i in 1:length(nodesize.values)) {
for (j in 1:length(mtry.values)) {
rf.tuned <- randomForest(Grad.Rate ~ ., data = CV_set, nodesize = nodesize.values[i], mtry = mtry.values[j])
oob.MSE[i, j] <- rf.tuned$mse[length(rf.tuned$mse)]
}
}
# Create a data frame for ggplot
plot.data <- expand.grid(nodesize = nodesize.values, mtry = mtry.values)
plot.data$OOB_MSE <- as.vector(oob.MSE)
# Plot using ggplot2 (faceted by mtry)
ggplot(plot.data, aes(x = nodesize, y = OOB_MSE)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ mtry, labeller = label_both) + # Facet by mtry
  labs(title = "OOB MSE vs. nodesize and mtry",
    x = "nodesize",
    y = "OOB MSE") +
  theme_bw()

#Finding which plot has the best parameters
best_params <- plot.data[which.min(plot.data$OOB_MSE), ]

bestRF <-randomForest(Grad.Rate~.,data = CV_set, nodesize = best_params$nodesize, mtry = best_params$mtry)
```

##### Boosting

```{r,echo=FALSE}
boost <- gbm(Grad.Rate~.,distribution = 'gaussian',data=CV_set,n.trees=500,interaction.depth=3,shrinkage=0.01,cv.folds = 10)

optNumIterations <- gbm.perf(boost, method = "cv")

boost_preds <- predict(boost, 
                       newdata = validation_set, 
                       n.trees = optNumIterations)

# Calculate your final Validation MSE for the boosting model
boost_validation_mse <- mean((validation_set$Grad.Rate - boost_preds)^2)
```

##### BART

```{r,echo=FALSE}
#predictors need to be separated
x_train <- model.matrix(Grad.Rate ~ ., data = CV_set)[, -1]
y_train <- CV_set$Grad.Rate # response

x_test <- model.matrix(Grad.Rate ~ ., data = validation_set)[, -1]
y_test <- validation_set$Grad.Rate

bart <- wbart(x.train = x_train, y.train = y_train, x.test = x_test)
```
